Semantics for the $ \lambda $-calculus

### 2024-04-30 report/chapters/category-theory.tex
Show that hat R is the Rezk completion of R

@@

\section{Adjunctions}

@@

@@

\begin{lemma}
  Every object $ c: \overline C $ is a retract of $ \iota_C(c_0) $ for some $ c_0: C $.
\end{lemma}
\begin{proof}
  Note that $ c = (c_0, a) $ for some $ c_0: C $ and an idempotent $ a: c \to c $. We have morphisms
  \begin{tikzcd}
    \iota_C(c) \arrow[r, shift left, "a_\rightarrow"] & (c, a) \arrow[l, shift left, "a_\leftarrow"]
  \end{tikzcd}, both given by $ a $. We have $ a_\leftarrow \cdot a_\rightarrow = a = \id{(c, a)} $, so $ (c, a) $ is a retract of $ \iota_C(c) $.
\end{proof}

@@

\begin{remark}
  Let $ D $ be a category. Suppose that we have a retraction-section pair in $ D $, given by
  \begin{tikzcd}
    d \arrow[r, shift left, "r"] & d^\prime \arrow[l, shift left, "s"]
  \end{tikzcd}.
  Now, suppose that we have an object $ c: D $ and a morphism $ f $ with $ (r \cdot s) \cdot f = f $. Then we get a morphism $ s \cdot f: d^\prime \to c $ such that $ f $ factors as $ r \cdot (s \cdot f) $. Also, for any $ g $ with $ r \cdot g = f $, we have
  \[ g = s \cdot r \cdot g = s \cdot f. \]
  \begin{center}
    \begin{tikzcd}
      d \arrow[r, left, "r"] \arrow[rd, "f"'] & d^\prime \arrow[r, "s"] \arrow[d, "s \cdot f" description] & d \arrow[ld, "f"]\\
      & c
    \end{tikzcd}.
  \end{center}
  Therefore, $ d^\prime $ is the equalizer of \begin{tikzcd}
    d \arrow[r, shift left, "\id d"] \arrow[r, shift right, "r \cdot s"'] & d
  \end{tikzcd}. In the same way, it is also the coequalizer of this diagram.

@@

>>>>

\section{Yoneda}
We can embed a category $ C $ into the functor category $ P C = [\op C, \SET] $ as follows (see \cite{Kashiwara}, Section 1.4):
\begin{definition}
  The \iindex{Yoneda embedding} $ Y : C \hookrightarrow P C $ is given on objects by $ Y(c) = C(-, c) $:
  \[ Y(c)(d) = C(d, c) \quad \text{and} \quad Y(c)(f)(g) = f \cdot g \]
  for $ d: C $, $ f: C(d, d^\prime) $ and $ g: C(d^\prime, c) $. It sends a morphism $ f: C(c, c^\prime) $ to the natural transformation $ Y(f): C(-, c) \Rightarrow C(-, c^\prime) $ given by
  \[ Y(f)(d)(g) = g \cdot f \]
  for $ d: C $ and $ g: C(d, c) $.
\end{definition}

For a functor between categories $ f: C \to D $, given the Yoneda embeddings $ Y_C : C \to P C $ and $ Y_D : D \to P D $ (we will often omit the subscript $ C $ and $ D $), we can create a diagram
\begin{center}
  \begin{tikzcd}
    C \arrow[r, "f"] \arrow[d, hookrightarrow, "Y"] & D \arrow[d, hookrightarrow, "Y"]\\
    P C & P D \arrow[l, "\op f_*"]
  \end{tikzcd}
\end{center}
Note that the arrows in this diagram are functors, so objects in a category, instead of elements of a set. Therefore, it does not make sense to talk about `equality' of the morphisms along the different paths, but we rather talk about isomorphism in the functor category $ [C, P C] $. If we have such an isomorphism, we say the diagram `2-commutes':
\begin{lemma}\label{lem:Yoneda-restriction-commutes}
  If $ f: C \to D $ is a fully faithful functor, the diagram above 2-commutes.
\end{lemma}
\begin{proof}
  For $ c, d : C $, since $ f $ is fully faithful, we have isomorphisms of sets, given by
  \[ Y(c)(d) = C(d, c) \xrightarrow[f_{d, c}]{\sim} D(f(d), f(c)) = \op f_*(Y(f(c)))(d). \]
  Also, for $ g: C(d^\prime, d) $, the following diagram commutes
  \begin{center}
    \begin{tikzcd}
      Y(c)(d) \arrow[d, "g \cdot -"] \arrow[r, "f_{d, c}"] & \op f_*(Y(f(c)))(d) \arrow[d, "f_{d^\prime, d}(g) \cdot -"]\\
      Y(c)(d^\prime) \arrow[r, "f_{d^\prime, c}"] & \op f_*(Y(f(c)))(d^\prime)
    \end{tikzcd}
  \end{center}
  so the isomorphism is natural in $ d $ and we have $ Y(c) \cong \op f_*(Y(f(c))) $ in $ P C $. Lastly, for $ g: C(c, c^\prime) $ and $ d: C $, the following diagram commutes
  \begin{center}
    \begin{tikzcd}
      Y(c)(d) \arrow[d, "- \cdot g"] \arrow[r, "f_{d, c}"] & \op f_*(Y(f(c)))(d) \arrow[d, "- \cdot f_{c, c^\prime}(g)"]\\
      Y(c^\prime)(d) \arrow[r, "f_{d, c^\prime}"] & \op f_*(Y(f(c^\prime)))(d)
    \end{tikzcd}
  \end{center}
  so the isomorphism is natural in $ c $ and we have $ Y \cong f \bullet Y \bullet \op f_* $ in $ [C, P C] $.
\end{proof}

@@

\section{Adjunctions and equivalences}

@@

\subsection{Weak equivalences}
There is also the notion of `weak equivalence'. In some cases, this is equivalent to an adjoint equivalence (for example, when its domain is univalent).
\begin{definition}
  A functor $ F: C \to D $ is called a \iindex{weak equivalence} if it is essentially surjective and fully faithful.
\end{definition}

@@

\begin{lemma}\label{lem:karoubi-is-retract}
  Every object $ c: \overline C $ is a retract of $ \iota_C(c_0) $ for some $ c_0: C $.
\end{lemma}
\begin{proof}
  Note that $ c = (c_0, a) $ for some $ c_0: C $ and an idempotent $ a: c \to c $. We have morphisms
  \begin{tikzcd}
    \iota_C(c) \arrow[r, shift left, "a_\rightarrow"] & (c, a) \arrow[l, shift left, "a_\leftarrow"]
  \end{tikzcd}, both given by $ a $. We have $ a_\leftarrow \cdot a_\rightarrow = a = \id{(c, a)} $, so $ (c, a) $ is a retract of $ \iota_C(c) $.
\end{proof}

@@

\begin{remark}\label{rem:retract-coequalizer}
  Let $ D $ be a category. Suppose that we have a retraction-section pair in $ D $, given by
  \begin{tikzcd}
    d \arrow[r, shift left, "r"] & d^\prime \arrow[l, shift left, "s"]
  \end{tikzcd}.
  Now, suppose that we have an object $ c: D $ and a morphism $ f $ with $ (r \cdot s) \cdot f = f $. Then we get a morphism $ s \cdot f: d^\prime \to c $ such that $ f $ factors as $ r \cdot (s \cdot f) $. Also, for any $ g $ with $ r \cdot g = f $, we have
  \[ g = s \cdot r \cdot g = s \cdot f. \]
  \begin{center}
    \begin{tikzcd}
      d \arrow[r, left, "r"] \arrow[rd, "f"'] & d^\prime \arrow[r, "s"] \arrow[d, "s \cdot f" description] & d \arrow[ld, "f"]\\
      & c
    \end{tikzcd}.
  \end{center}
  Therefore, $ d^\prime $ is the equalizer of \begin{tikzcd}
    d \arrow[r, shift left, "\id d"] \arrow[r, shift right, "r \cdot s"'] & d
  \end{tikzcd}. In the same way, it is also the coequalizer of this diagram.

@@

Now, there is also another (classically equivalent) definition of the category of retracts:
\begin{definition}
  Let $ Y: C \to P C $ be the Yoneda embedding of $ C $ into its category of presheaves. We define the category $ \hat \R $ as the full subcategory of $ P C $ consisting of objects $ F : P C $ such that there exists an object $ c: C $ and a retraction of $ Y(c) $ onto $ F $. Note that the existence of $ c $ and the retraction is \textit{mere existence} (see the chapter on Univalent Foundations) because we need this to be a subcategory of $ P C $.
\end{definition}

\begin{lemma}
  $ \hat \R $ is the Rezk completion of $ \R $.
\end{lemma}
\begin{proof}
  First of all, note that $ \hat \R $ is univalent, since it is a presheaf category (see Theorem 4.5 in \cite{univalent-categories}). So we must show that we have a weak equivalence from $ \R $ to $ \hat \R $.

  Secondly, by Corollary \ref{cor:karoubi-presheaf} we have an adjoint equivalence $ {\op{\iota_C}}_*: P \R \xrightarrow \sim P C $ and by Lemma \ref{lem:Yoneda-restriction-commutes}, the following diagram 2-commutes:
  \begin{center}
    \begin{tikzcd}
      C \arrow[r, "\iota_C"] \arrow[d, "Y"] & \R \arrow[d, "Y"]\\
      P C & P \R \arrow[l, "\sim"', "{\op{\iota_C}}_*"]
    \end{tikzcd}
  \end{center}

  By Lemma \ref{lem:karoubi-is-retract}, every object of $ \R $ is a retract of some object in $ c $. Therefore, every object in the image of $ \R $ inside $ P C $ is a retract of some object in the image of $ C $, so we can turn $ Y \bullet {\op{\iota_C}}_* : \R \to P C $ into a fully faithful embedding into $ \hat \R $.

  Now, to show that it is essentially surjective, take an object $ F : \hat \R $. That is, take a retract \begin{tikzcd} F \arrow[r, shift left, "s"] & Y(c) \arrow[l, shift left, "r"] \end{tikzcd}. Note that $ r \cdot s $ is an idempotent on $ Y(c) $. Since $ Y $ is a fully faithful embedding, it gives an equivalence on the hom-sets, so we have an element $ e : C(c, c) $ that maps to $ r \cdot s : P C (Y(c), Y(c)) $. We will show that $ (c, e) $ is our preimage in $ \R $ of $ F $.

  Also, $ (c, e) $ is a retract of $ \iota_C(c) $ with section and retraction $ e $. Therefore, $ \op{\iota_C} \bullet Y(c, e) $ is a retract of $ \op{\iota_C} \bullet Y (\iota_C (c)) $ with section and retraction $ \op{\iota_C} \bullet Y(e) $. Composing these gives an idempotent
  \[ \op{\iota_C} \bullet Y(e): d \mapsto f \mapsto f \cdot e \]
  on $ \op{\iota_C} \bullet Y (\iota_C (c)) $. Transporting this along the isomorphism with $ Y(c) $, we get an idempotent
  \[ d \mapsto f \mapsto \iota_C^{-1}(\iota_C(f) \cdot e) \]
  on $ Y(c) $. Note that we can view $ e $ both as an idempotent on $ c $ and on $ \iota_C(e) $, and that $ \iota_C(e) = e $, so $ \iota_C^{-1}(\iota_C(f) \cdot e) = f \cdot e $.

  Therefore, the idempotent on $ Y(c) $ is exactly $ Y(e) = r \cdot s $. By Remark \ref{rem:retract-coequalizer}, $ \op{\iota_C} \bullet Y(c, e) $ is the equalizer of $ \id{Y(c)} $ and $ r \cdot s $. Note that by the same remark, $ F $ is the equalizer of $ \id{Y(c)} $ and $ r \cdot s $ as well. Since the equalizer is unique up to unique isomorphism, we have an isomorphism $ F \cong \op{\iota_C} \bullet Y(c, e) $ and we see that $ Y \bullet {\op{\iota_C}}_* $ is essentially surjective.
\end{proof}

\begin{remark}
  As shown above, $ \R $ and $ \hat \R $ have a very strong relation. In classical mathematics, they are even equivalent. However, since $ \R $ is not univalent, we do not have a full equivalence in univalent foundations and our choice of definition will have consequences. One one hand, $ \hat \R $ is univalent. On the other hand, the objects of $ \R $ are uniquely associated with one idempotent morphism in $ C $.

  In this thesis, I will choose to interpret the `category of retracts' to mean $ \R $. This is because Dana Scott and Paul Taylor perform most of their constructions explicitly using the (idempotent) morphisms of $ C $, which is closest to working in $ \R $.

  It is however very well possible that most of the proofs can be made to work in $ \hat \R $ as well. For future research, it would be interesting to see what the differences are between working in $ \R $ and $ \hat \R $ for these proofs.
\end{remark}

### 2024-04-30 report/chapters/paper.tex
Show that hat R is the Rezk completion of R

Note that if we consider $ L_1 $ as a monoid, with operation $ \bullet $ and unit $ x_1 $, and if we consider this monoid as a category, we can embed this (fully faithfully) into $ \mathbf L $. This gives the following sequence of embeddings:
\begin{center}
  \begin{tikzcd}
    L_1 \arrow[r, hookrightarrow] & \mathbf L \arrow[r, hookrightarrow] & \R \arrow[r, hookrightarrow] & \Pshf L
  \end{tikzcd}
\end{center}
Note that the composition of the first two morphisms sends the object of the category $ L_1 $ to $ (\lambda x_1, x_1) : \R $, which is exactly the usual embedding of $ L_1 $ into its Karoubi envelope.

Write $ i: \op{L_1} \hookrightarrow \op{\mathbf L} $ and $ j: \op{\mathbf L} \hookrightarrow \op \R $ for the embeddings on the opposite categories. By Corollary \ref{cor:surjective-precomposition}, the first three categories in this sequence yield the following sequence of essentially surjective functors on the presheaf categories:
\begin{center}
  \begin{tikzcd}
    & \Pshf L \arrow[d, "\sim"]\\
    P \R \arrow[r, twoheadrightarrow, "j_*"] \arrow[rr, bend right, "\sim"] & P \mathbf L \arrow[r, twoheadrightarrow, "i_*"] & P L_1
  \end{tikzcd}
\end{center}
The two equivalences in this diagram are due to Lemma \ref{lem:lawvere-presheaf} and Corollary \ref{cor:karoubi-presheaf}.

@@

@@

@@

The fundamental theorem states that there is an equivalence of categories
\[ \LamTh \cong \Alg \Lambda. \]
We will prove this by showing that there is a fully faithful and essentially surjective functor.

>>>>

Note that if we consider $ L_1 $ as a monoid, with operation $ \bullet $ and unit $ x_1 $, and if we consider this monoid as a category, we can embed this (fully faithfully) into $ \mathbf L $. This gives the following sequence of embeddings:
\begin{center}
  \begin{tikzcd}
    L_1 \arrow[r, hookrightarrow] & \mathbf L \arrow[r, hookrightarrow] & \R \arrow[r, hookrightarrow] & \Pshf L
  \end{tikzcd}
\end{center}

Note that the composition of the first two morphisms sends the object of the category $ L_1 $ to $ (\lambda x_1, x_1) : \R $, which is exactly the usual embedding of $ L_1 $ into its Karoubi envelope.

Write $ i: \op{L_1} \hookrightarrow \op{\mathbf L} $ and $ j: \op{\mathbf L} \hookrightarrow \op \R $ for the embeddings on the opposite categories. By Corollary \ref{cor:surjective-precomposition}, the first two morphisms in this sequence yield the following sequence of essentially surjective functors on the presheaf categories:
\begin{center}
  \begin{tikzcd}
    \mathbf R \arrow[r, hookrightarrow] \arrow[d, hookrightarrow, "Y"] & \Pshf L \arrow[d, "\sim"]\\
    P \R \arrow[r, twoheadrightarrow, "j_*"] \arrow[rr, bend right, "\sim"] & P \mathbf L \arrow[r, twoheadrightarrow, "i_*"] & P L_1
  \end{tikzcd}
\end{center}
The two equivalences in this diagram are due to Lemma \ref{lem:lawvere-presheaf} and Corollary \ref{cor:karoubi-presheaf}.

@@

\begin{lemma}
  We have a natural isomorphism between the two functors from $ \R $ to $ P \mathbf L $ in the diagram above.
\end{lemma}
\begin{proof}
  Note:
  \[ (Y \bullet j_*)(A)(n) = (j \bullet (Y(A)))(n) = Y(A)(j(n)) = \R(U^n, A) \]
  \[ (\iota \bullet \sim)(A)(n) = \{ a : L_n \mid A a = a \} \]

  By Scott's representation theorem, we have $ \mathbb R(U^n, U) \cong L_n, $ via
  \[ f \mapsto \iota_{0, n}(f) ((((c, x_1), x_2), \dots), x_n). \]
  Restricting this equivalence to the morphisms to $ A $ yields
  \[ \R(U^n, A) \cong \{ a : L_n \mid A a = a \}. \]
  \TODO: Check coherence.
\end{proof}

@@

Recall that Paul Taylor constructed a class of display maps for $ \R $.

@@

The fundamental theorem states that there is an adjoint equivalence
\[ \LamTh \cong \Alg \Lambda. \]
We will prove this by showing that there is a weak equivalence.

### 2024-04-30 report/chapters/previous-work.tex
Show that hat R is the Rezk completion of R

  However, note that we have a weak equivalence (an essentially surjective and fully faithful functor) between strict categories (categories in which the type of objects is a homotopy set) $ \sum_A: \R^A \xrightarrow \sim (\R \downarrow_D A) $. If we assume the axiom of choice, $ \sum_A $ becomes an equivalence of categories (\cite{univalent-categories}, introduction, page 2 \TODO: maybe find a better reference for this) and then our pullback functor is given by
  \[ \sum_A^{-1} \bullet (\alpha \cdot -) \bullet \sum_B, \]
  where $ (\alpha \cdot -): \R^A \to \R^B $ denotes the functor given by precomposition with $ \alpha $.
\end{remark}

>>>>

  However, note that we have a weak equivalence between strict categories (categories in which the type of objects is a homotopy set) $ \sum_A: \R^A \xrightarrow \sim (\R \downarrow_D A) $. If we assume the axiom of choice, $ \sum_A $ becomes an equivalence of categories (\cite{univalent-categories}, introduction, page 2 \TODO: maybe find a better reference for this) and then our pullback functor is given by
  \[ \sum_A^{-1} \bullet (\alpha \cdot -) \bullet \sum_B, \]
  where $ (\alpha \cdot -): \R^A \to \R^B $ denotes the functor given by precomposition with $ \alpha $.
\end{remark}

### 2024-04-30 report/chapters/univalent-foundations.tex
Show that hat R is the Rezk completion of R

@@

Lastly, a result about univalent categories that we will use a couple of times in this thesis:
\begin{lemma}
  For a functor between univalent categories $ F: A \to B $, the types `$ F $ is an adjoint equivalence' and `$ F $ is fully faithful and essentially surjective' are equivalent propositions (see \ref{sec:homotopy-props-sets}).
\end{lemma}
\begin{proof}
  See \cite{univalent-categories}, Lemma 6.8.
\end{proof}

>>>>

Note that if $ B $ is a univalent category and $ A $ is any cateory, the functor category $ [A, B] = B^A $ is univalent as well. In particular, the presheaf category $ P A = [\op A, \SET] $ is univalent.

\begin{lemma}
  For every category $ C $, there exists a univalent category $ \hat C $ with a weak equivalence
  \[ \iota: C \hookrightarrow \hat C. \]
\end{lemma}
\begin{proof}
  See \cite{univalent-categories}, Theorem 8.5.
\end{proof}
We call this univalent category the \iindex{Rezk completion}.
\TODO: Uniqueness?

@@

Lastly, a result about univalent categories that we will use a couple of times in this thesis:
\begin{lemma}
  For a functor between univalent categories $ F: A \to B $, the types `$ F $ is an adjoint equivalence' and `$ F $ is a weak equivalence' are equivalent propositions (see \ref{sec:homotopy-props-sets}).
\end{lemma}
\begin{proof}
  See \cite{univalent-categories}, Lemma 6.8.
\end{proof}
